Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using Multiple Automatically Generated Templates

### 2013-09-01 paper.Rnw
Added ADNI boilerplate

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
Multiple Automatically Generated Templates}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,8]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,8]{Aristotle N. Voineskos}
\author[1,6,7]{M. Mallar Chakravarty}
\author[ ]{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[8]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
neuroscientists because it is implicated in many forms of brain disfunction,
including Alzheimer's disease \citep{Sabuncu2011} and schizophrenia
\citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies, structural magnetic
resonance images (MRI) are often used for the volumetric assessment of the
hippocampus.  As such, accurate segmentation of the hippocampus and its
subregions in MRI is a necessary first step to better understand the
inter-individual variability of subject neuroanatomy. Typically, the gold
standard for neuroanatomical segmentation is manual delineation by an expert
human rater.  However, with the availability of increasingly large MRI datasets,
the time and expertise required for manual segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. Further, there is
significant variation between segmentation protocols regarding specific
anatomical boundaries of the hippocampus \citep{Geuze2004} and this has led to
efforts to create an unified hippocampal segmentation protocol \citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
can be considered complementary to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to inaccuracies in the estimated nonlinear transformation itself,
partial volume effects in label resampling, and irreconcilable differences
between the neuroanatomy represented within the atlas and target images.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm \citep{Warfield2004} that computes a
probabilistic segmentation using an expectation maximization framework from an
set of competing segmentations, or others where a subset of segmentations can be
estimated using metrics such as the sum of squared differences in the regions of
interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our multi-atlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

Of special relevance to the present work is the LEAP algorithm \citep{Wolz2010}
because of its focus on performing multi-atlas segmentation with a limited
number of input atlases.  The LEAP algorithm is a clever modification to the
basic multi-atlas strategy in which an atlas library is grown, beginning with a
set of manually labelled atlases, by successively incorporating unlabelled
target images once they themselves have been labelled using multi-atlas
techniques. The sequence in which target images are labelled is chosen so that
the similarity between the atlas images and the target images is minimised at
each step, effectively allowing for deformations between very dissimilar images
to be broken up into sequences of smaller deformations. Although
\citet{Wolz2010} begin with an atlas library of 30 MR images, this method could
theoretically work using a much smaller atlas library.  In their validation,
LEAP was used to segment the whole hippocampus in the ADNI-1 baseline dataset,
achieving a mean Dice score of 0.85 against \gs segmentations.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in
other publications \citep{Collins2010,Heckemann2011,Aljabar2009}. When only a
single atlas is used, basic multi-atlas segmentation degenerates into
model-based segmentation: labels are propogated from the atlas to a target, and
no label fusion is needed.

\mb bootstraps the creation of a large template library given a limited input atlas 
library, and then uses the template library in basic multi-atlas segmentation. 
Images for the template library are selected from a set of input target images, 
either arbitrarily or so as to reflect the neuroanatomy or demographics of the 
target set as a whole (for instance, by sampling equally from cases and controls). 
The template library images are automatically labelled by each of the atlases
via label propagation. Basic multi-atlas segmentation is then conducted using
the template library to segment the entire set of target images (including the
targets whose images are used in the construction of the template library).
Since each template library image has multiple labels (one from each atlas), the
final number of labels to be fused for each target may be quite large (i.e. \#
of atlas $\times$ \# of templates).

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

\paragraph{\adnidataset dataset} 
69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we will use these segmentations as the \gs for comparison.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

>>>>

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
Multiple Automatically Generated Templates}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,8]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,8]{Aristotle N. Voineskos}
\author[1,6,7]{M. Mallar Chakravarty}
\author[ ]{and the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[8]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
neuroscientists because it is implicated in many forms of brain disfunction,
including Alzheimer's disease \citep{Sabuncu2011} and schizophrenia
\citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies, structural magnetic
resonance images (MRI) are often used for the volumetric assessment of the
hippocampus.  As such, accurate segmentation of the hippocampus and its
subregions in MRI is a necessary first step to better understand the
inter-individual variability of subject neuroanatomy.

Typically, the gold standard for neuroanatomical segmentation is manual
delineation by an expert human rater.  However, with the availability of
increasingly large MRI datasets, the time and expertise required for manual
segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. Further, there is
significant variation between segmentation protocols regarding specific
anatomical boundaries of the hippocampus \citep{Geuze2004} and this has led to
efforts to create an unified hippocampal segmentation protocol \citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
can be considered complementary to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to error in the estimated nonlinear transformation itself, partial
volume effects in label resampling, and irreconcilable differences between the
neuroanatomy represented within the atlas and target images.

@@

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm (Simultaneous Truth And Performance Level
Estimation; \citet{Warfield2004}) that computes a probabilistic segmentation
using an expectation-maximization framework from an set of competing
segmentations, or others where a subset of segmentations can be estimated using
metrics such as the sum of squared differences in the regions of interest to be
segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our multi-atlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

Of special relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing multi-atlas segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic multi-atlas strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using multi-atlas techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs segmentations.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic multi-atlas segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic multi-atlas segmentation degenerates into model-based segmentation: labels
are propogated from the atlas to a target, and no label fusion is needed.

\mb ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic multi-atlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation.  Basic multi-atlas
segmentation is then conducted using the template library to segment the entire
set of target images (including the target images used in the construction of
the template library).  Since each template library image has multiple labels
(one from each atlas), the final number of labels to be fused for each target
may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in Experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

\paragraph{\adnidataset dataset} 
Data used in the preparation of this article were obtained from the Alzheimer’s
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a $60 million, 5-year public- private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer’s disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California – San Francisco. ADNI is the result
of efforts of many co- investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2.  To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal older individuals, people with early or late
MCI, and people with early AD. The follow up duration of each group is specified
in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally recruited
for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For up-to-date
information, see www.adni-info.org.

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we will use these segmentations as the \gs for comparison.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.
